---
title: "Introduction to the mtarm Package"
author: 
- "Luis Hernando Vanegas"
- "Sergio Alejandro Calderón"
- "Luz Marina Rondón"

output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true

bibliography: references.bib

vignette: >
  %\VignetteIndexEntry{Introduction to the mtarm Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Multivariate Threshold Autoregressive (TAR) models

The `mtarm` package provides a computational tool designed for Bayesian estimation, inference, and forecasting in multivariate Threshold Autoregressive (TAR) models. These models provide a versatile approach for modeling nonlinear multivariate time series and include multivariate Self-Exciting Threshold Autoregressive (SETAR) and Vector Autoregressive (VAR) models as particular cases [@VANEGAS2025]. The package accommodates a broad class of innovation distributions beyond the Gaussian assumption, such as Student-$t$, slash, symmetric hyperbolic, Laplace, contaminated normal, skew-normal, and skew-$t$ distributions, thereby enabling robust modeling of heavy tails, asymmetry, and other non-Gaussian characteristics.

## Installation

### Install from GitHub

```{r eval = FALSE}
remotes::install_github("lhvanegasp/mtarm")
```

### Install from CRAN

```{r eval = FALSE}
install.packages("mtarm")
```

## Application: Temperature, precipitation, and two river flows in Iceland

### Dataset

The data are available in the object \`iceland.rf\` and were obtained from [@T90], who provided a detailed description of the geographical and meteorological characteristics of the rivers and analyzed each series individually. Subsequently, [@T98] conducted a bivariate analysis of the same dataset. The focus is on the bivariate time series $\{(Y_{1,t},Y_{2,t})^{\top}\}_{t\geq 1}$, where $Y_{1,t}$ and $Y_{2,t}$ denote the daily river flow (in cubic meters per second, ${m}^3/{s}$) of the Jökulsá Eystri and Vatnsdalsá rivers, respectively. The sample covers the period from 1972 to 1974, comprising 1095 observations. The exogenous variables include daily precipitation $X_t$, measured in millimeters (${mm}$), and temperature $Z_t$, measured in degrees Celsius ($^\circ\mathrm{C}$), both recorded at the meteorological station in Hveravellir. Precipitation corresponds to the accumulated rainfall from 9:00 A.M. of the previous day to 9:00 A.M. of the current day.

```{r fig.width=9, fig.height=7}
library(mtarm)
data(iceland.rf)       
str(iceland.rf)     
```

```{r fig.width=9, fig.height=5}
summary(iceland.rf[,-5])
```

```{r fig.width=9, fig.height=5}
plot(ts(as.matrix(iceland.rf[,-5])), main="Iceland")
```

### Model specification

Following [@T98], the series are modeled using a $\mathrm{TAR}(2; p=(15,15), q=(4,4), d=(2,2))$ specification given by

\$\$Y_t=\\sum\\limits\_{j=1}\^2 I(Z\_{t-h}\\in(c\_{j-1},c_j])\\Big(\\!{\\phi}\_0\^{\^{(j)}}+\\sum\\limits\_{i=1}\^{15}\\boldsymbol{\\phi}\_i\^{\^{(j)}}Y\_{t-i}+\\sum\\limits\_{i=1}\^{4}\\boldsymbol{\\beta}\_i\^{\^{(j)}}\\!X\_{t-i}+\\sum\\limits\_{i=1}\^{2}{\\delta}\_i\^{\^{(j)}}\\!Z\_{t-i}+\\epsilon_t\^{\^{(j)}}\\!\\Big),\$\$

where $\epsilon_t^{^{(j)}}$ is the error term. The last 55 observations (from November 7 to December 31, 1974), corresponding to $5\%$ of the sample, are excluded from the estimation stage and reserved for out-of-sample forecast evaluation. The following code requests the estimation for the $\mathrm{TAR}(2; p=(15,15), q=(4,4), d=(2,2))$ specification under Gaussian, Slash, Student-$t$, and Laplace error distributions.

### Parameter estimation

```{r}
fits <- mtar_grid(~ Jokulsa + Vatnsdalsa | Temperature | Precipitation,
                  data=iceland.rf, subset={Date<="1974-11-06"}, 					      
                  row.names=Date, nregim.min=2, nregim.max=2, p.min=15,  				
                  p.max=15, q.min=4, q.max=4, d.min=2, d.max=2,  					      
                  n.burnin=200, n.sim=300, n.thin=3, ssvs=TRUE,
                  dist=c("Gaussian","Slash","Student-t","Laplace"),
                  plan_strategy="multisession")

fits
```

### Model comparison using forecast accuracy measures

#### Adjusted within-sample

The following code requests Deviance Information Criterion (DIC) [@spiegelhalter02; @spiegelhalter14] and Watanabe-Akaike Information Criterion (WAIC) [@w10] values.

```{r}
DICs <- DIC(fits)
DICs

WAICs <- WAIC(fits)
WAICs
```

#### Out-of-sample

In addition, the following code provides the median of the log-score [@Good1952], the Energy Score (ES) [@GneitingEtAl2008TEST]—a multivariate extension of the Continuous Ranked Probability Score (CRPS)[@MathesonWinkler1976; @GrimitGneitingBerrocalJohnson2006]—and the Absolute Percentage Error (APE), all computed from the observed and forecasted values for the last 55 observations.

```{r}
newdata <- subset(iceland.rf, Date>"1974-11-06") 
oos <- out_of_sample(fits, newdata=newdata, n.ahead=nrow(newdata),  
                     by.component=TRUE, FUN=median) 
oos[,c(1,2,5,6)]
```

### Residuals

```{r fig.width=9, fig.height=5}
res <- residuals(fits$`Student-t.2.15.4.2`)  
```

```{r fig.width=9, fig.height=5}
par(mfrow=c(1,2)) 
qqnorm(res$full, pch=20, col="blue", main="") 
abline(0, 1, lty=3) 
hist(res$full, freq=FALSE, xlab="Quantile-type residual", ylab="Density", main="") 
curve(dnorm(x), col="blue", add=TRUE)
```

```{r fig.width=9, fig.height=5}
par(mfrow=c(1,2))  
acf(res$full, col="blue", main="")
pacf(res$full, col="blue", main="")
```

### Forecasting

```{r}
pred <- predict(fits$`Student-t.2.15.4.2`, newdata=newdata, n.ahead=nrow(newdata),
                row.names=Date, credible=0.8)

head(pred$summary)
tail(pred$summary)
```

### Convergence diagnostics

#### Geweke statistic

```{r}
geweke_diagTAR(fits$`Student-t.2.15.4.2`)

```

#### Effective sample size

```{r}
effectiveSize_TAR(fits$`Student-t.2.15.4.2`) 
```
